---
title: "Estimating between-species variance"
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mvtnorm)
library(tidyverse)
```

## Parameters 

```{r parameters}
n = 4 # n per species
N = 7 # number of species

bVar = 1 # between species var
wVar = 1 # within species var

K = 2000 # number of "genes"
```

## Simulate 

```{r generate}
# generate a multivariate distribution
iExp = rep(1:N,each=n)
sigma = diag(rep(bVar,N))[iExp,iExp] + diag(rep(wVar,N*n))
dat <- rmvnorm(K,sigma = sigma)
```

```{r}
image(sigma,col = colorRampPalette(c("white","red"))(100))
```


## Estimate variances, method 1

* Between-species variance is esitmated as the variance between species means.
* Within-species variance is esitmated as the mean of the species variances.

```{r estimate1}
# estimate parameters using var and mean
colSpeciesIndices <- split(1:length(iExp), f = iExp)
species.mean <- sapply(colSpeciesIndices, function(i){ rowMeans(dat[ ,i]) })
species.var <- sapply(colSpeciesIndices, function(i){ apply(dat[ ,i],1,var) })
bVarEst1 <- apply(species.mean,1,var)
wVarEst1 <- rowMeans(species.var)
```

```{r plotEst1}
qplot(bVarEst1,wVarEst1) + coord_fixed() +
  geom_hline(yintercept = wVar, col="red") + 
  geom_vline(xintercept = bVar, col="red") +
  geom_smooth(method="lm")

```

## Estimate variances, method 2

Use maximum likelihood to fit the data to multivariate normal distribution.

```{r estimate2}
# estimate using ML on mvnorm
lapply(1:nrow(dat), function(i){
  optim(par = log(c(10,10)), method = "L-BFGS-B", datRow = dat[i, ],
        fn = function(par, datRow){
          bVarEst = exp(par[1]); wVarEst = exp(par[2])
          sigma = diag(rep(bVarEst,N))[iExp,iExp] + diag(rep(wVarEst,N*n))
          ll = dmvnorm(x = datRow,sigma = sigma,log = T)  
          # cat(sprintf("bVarEst:% 10.4f  wVarEst:% 10.4f  ll:% 10.4f\n",bVarEst,wVarEst,ll))
          return(-ll)
  })
}) -> res

bVarEst2 <- sapply(res, function(r) exp(r$par[1])) 
wVarEst2 <- sapply(res, function(r) exp(r$par[2]))
```

```{r plotEstimate2}
qplot(bVarEst2,wVarEst2) + coord_fixed() + 
  geom_hline(yintercept = wVar, col="red") + 
  geom_vline(xintercept = bVar, col="red") +
  geom_smooth(method="lm")
```

## compare

```{r plotCompare}
qplot(bVarEst1,bVarEst2) + geom_abline(slope=1, intercept = 0, col="red")
qplot(wVarEst1,wVarEst2) + geom_abline(slope=1, intercept = 0, col="red")
```

```{r plotComparebVarDist}
tibble(bVarEst1,bVarEst2) %>% 
  gather() %>% 
  ggplot() +
  geom_density( aes(x=value, color=key))
```

* Mean squared error for bVarEst1: `r mean((bVarEst1-bVar)^2)`
* Mean squared error for bVarEst2: `r mean((bVarEst2-bVar)^2)`

