---
title: "Comparison with original EVE results (simData)"
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(dev = "svg")
library(ape)

source('./scripts/eve-io.R')
source('./scripts/dvdt.R')
source('./scripts/twoThetaTest.R')

library(tidyverse)
```

```{r Load data}
# simulated example data
tree <- read.tree("data/comparison/newFormat/simData/examplePhyloNamed.newick")
gene.data <- getExprData("data/comparison/newFormat/simData/sampleExpr.tsv")

# original EVE results
indivBetaMLparams <- read.table("data/comparison/oldEVE/results/indivBetaMLparams_simData.res", 
                                col.names = c("theta","sigma2","alpha","beta"))
sharedBetaMLparams <- read.table("data/comparison/oldEVE/results/sharedBetaMLparams_simData.res", 
                                 col.names = c("theta","sigma2","alpha","beta"))
betaTestLRTs <- scan("data/comparison/oldEVE/results/betaTestLRTs_simData.res")
sharedBetaMLs <- scan("data/comparison/oldEVE/results/sharedBetaMLs_simData.res")
indivBetaMLs <- scan("data/comparison/oldEVE/results/indivBetaMLs_simData.res")

# original EVE two-theta test results
oneThetaMLparams <- read.table("data/comparison/oldEVE/results/oneThetaMLparams_simData.res", 
                                col.names = c("theta","sigma2","alpha","beta"))
twoBSThetaMLparams <- 
  read.table("data/comparison/oldEVE/results/twoBSThetaMLparams_simData.res", 
             col.names = c("theta","sigma2","alpha","beta")) %>% 
  mutate(key=paste0("theta",rep(1:2,n()/2))) %>% 
  spread(key = key,value = theta)
BSThetaTestLRTs <- scan("data/comparison/oldEVE/results/BSThetaTestLRTs_simData.res")
oneThetaMLs <- scan("data/comparison/oldEVE/results/oneThetaMLs_simData.res")
twoBSThetaMLs <- scan("data/comparison/oldEVE/results/twoBSThetaMLs_simData.res")
```

## Beta shared test

```{r Run beta-shared test, cache=TRUE, include=F}
BetaSharedRes <- betaSharedTest(tree, gene.data)
```

```{r defPlotFun}
myCompareParamPlot <- function(df){
  df %>% 
    mutate( Parameter = factor(Parameter,levels=unique(Parameter))) %>% 
    mutate( geneID = factor(geneID,levels=rev(unique(geneID)))) %>% 
    ggplot( aes(x=Value,y=geneID, color=implementation, shape=implementation)) + 
    geom_point( size=3) + 
    scale_shape_manual(values = c(3,4)) +
    scale_x_log10() +
    facet_grid(. ~ Parameter ,scales = "free") +
    theme(legend.position="bottom")
}
```

### Beta shared estimate

* old: `r sharedBetaMLparams$beta[1]`
* new: `r BetaSharedRes$sharedBeta`

### LRTs

```{r plotLRT, fig.height=3, message=FALSE}
bind_rows( .id="implementation",
  old=tibble( LRT=betaTestLRTs, geneID=rownames(gene.data)),
  new=tibble( LRT=BetaSharedRes$LRT, geneID=rownames(gene.data))) %>% 
  gather(key = "Parameter",value = "Value", -implementation, -geneID) %>%
  myCompareParamPlot() +
  geom_vline(data=tibble(x=qchisq(p=0.95,df=1),Parameter="LRT"),aes(xintercept=x),color="blue",linetype=3) +
  geom_vline(data=tibble(x=qchisq(p=0.05,df=1),Parameter="LRT"),aes(xintercept=x),color="blue",linetype=3) +
  scale_x_continuous()
```

Vertical lines are 0.05 and 0.95 of chi-squared distribution with 1 degree of freedom.

### MLs

```{r plotMLs, fig.height=3}
bind_rows( .id="implementation",
  old=tibble( indivBetaML=indivBetaMLs,sharedBetaML=sharedBetaMLs, geneID=rownames(gene.data)),
  new=tibble( indivBetaML=-map_dbl(BetaSharedRes$indivBetaRes,"value"),sharedBetaML=-map_dbl(BetaSharedRes$sharedBetaRes,"value"), geneID=rownames(gene.data))) %>% 
  gather(key = "Parameter",value = "Value", -implementation, -geneID) %>%
  mutate( geneID = factor(geneID,levels=rev(unique(geneID)))) %>% 
  ggplot( aes(x=Value,y=geneID, color=implementation, shape=implementation)) + 
  geom_point( size=3) + 
  scale_shape_manual(values = c(3,4)) +
  # scale_x_log10() +
  facet_grid(. ~ Parameter ,scales = "free") +
  theme(legend.position="bottom")
```


Note the one gene (GENE7) that has positive values!? This should not be possible as this are log likelihoods, i.e. positive values means that the likelihood is larger than 1.


### individual beta parameters

```{r plotCompareIndivBetaParams, fig.height=3}
bind_rows(.id = "implementation",
          new=map_df(BetaSharedRes$indivBetaRes, ~ as_tibble(t(.x$par))) %>% 
            mutate( geneID = rownames(gene.data)),
          old=indivBetaMLparams %>% 
            mutate( geneID = rownames(gene.data))) %>% 
  gather(key = "Parameter",value = "Value", -implementation, -geneID) %>%
  myCompareParamPlot() +
  geom_vline(data=tibble(x=46.0517,Parameter="alpha"),aes(xintercept=x),color="red",linetype=2)
```

Five of the 12 genes have significantly different parameter estimates and all of these seem to bump up against the upper bound for alpha. Note that GENE1 has an extremely high beta in the old implementation.

### without alpha bounds

```{r fitIndivBetaNoBound}
fitIndivBetaNoBound <- function(tree, gene.data, colSpecies = colnames(gene.data), alphaMax=Inf){
  #Calculate the per gene parameter matrix based on the gene data
  initial.param.matrix <- initialParams(gene.data, colSpecies)
  
  # match the column species with the phylogeny tip labels
  index.expand <- match(colSpecies, tree$tip.label)

  # For each gene, optimize the parameters and store the resulting likelihood in the likelihood vector
  lapply(1:nrow(gene.data), function(row){
    # Error handling to catch infinte optim or function values that arise when data with NaN paramters is optimized
    res <- tryCatch({
      optim(initial.param.matrix[row, ], fn = LLPerGeneIndivBeta, gr = NULL, tree, gene.data[row, ], index.expand,
            method = "L-BFGS-B", lower = c(-Inf, 1e-10, 1e-10, 1e-10), upper = c(Inf, Inf, alphaMax, Inf))
    }, error = function(e) {
      warning(paste(e$message, "at gene.data row", row), immediate. = T)
    })
  }) -> res
  
  return(res)
}

indivBetaResNoBound <- fitIndivBetaNoBound(tree,gene.data)
```

```{r plotCompareIndivBetaParamsNoBoundAlpha, fig.height=3}
bind_rows(.id = "implementation",
          "new (no bound)"=map_df(indivBetaResNoBound, ~ as_tibble(t(.x$par))) %>% 
            mutate( geneID = rownames(gene.data)),
          old=indivBetaMLparams %>% 
            mutate( geneID = rownames(gene.data))) %>% 
  gather(key = "Parameter",value = "Value", -implementation, -geneID) %>%
  myCompareParamPlot()
```

### beta shared estimates

```{r plotCompareSharedBetaParams, fig.height=3}
bind_rows(.id = "implementation",
          new=map_df(BetaSharedRes$sharedBetaRes, ~ as_tibble(t(.x$par))) %>% 
            mutate( geneID = rownames(gene.data)),
          old=sharedBetaMLparams %>% 
            mutate( geneID = rownames(gene.data))) %>% 
  gather(key = "Parameter",value = "Value", -implementation, -geneID) %>%
  filter( Parameter != "beta") %>% 
  myCompareParamPlot() +
  geom_vline(data=tibble(x=46.0517,Parameter="alpha"),aes(xintercept=x),color="red",linetype=2)

```

## Two-theta test results

```{r Run two-theta test, cache=TRUE, include=F}
twoThetaRes <- twoThetaTest(tree, gene.data,shiftSpecies = c("D","E"))
```

### LRTs

```{r plotLRTtwoTheta, fig.height=3, message=FALSE}
bind_rows( .id="implementation",
  old=tibble( LRT=BSThetaTestLRTs, geneID=rownames(gene.data)),
  new=tibble( LRT=twoThetaRes$LRT, geneID=rownames(gene.data))) %>% 
  gather(key = "Parameter",value = "Value", -implementation, -geneID) %>%
  myCompareParamPlot() +
  geom_vline(data=tibble(x=qchisq(p=0.95,df=1),Parameter="LRT"),aes(xintercept=x),color="blue",linetype=3) +
  geom_vline(data=tibble(x=qchisq(p=0.05,df=1),Parameter="LRT"),aes(xintercept=x),color="blue",linetype=3) +
  scale_x_continuous()
```

### MLs

```{r plotMLstwoTheta, fig.height=3}
bind_rows( .id="implementation",
  old=tibble( oneThetaML=oneThetaMLs,twoThetaML=twoBSThetaMLs, geneID=rownames(gene.data)),
  new=tibble( oneThetaML=-map_dbl(twoThetaRes$oneThetaRes,"value"),twoThetaML=-map_dbl(twoThetaRes$twoThetaRes,"value"), geneID=rownames(gene.data))) %>% 
  gather(key = "Parameter",value = "Value", -implementation, -geneID) %>%
  mutate( geneID = factor(geneID,levels=rev(unique(geneID)))) %>% 
  ggplot( aes(x=Value,y=geneID, color=implementation, shape=implementation)) + 
  geom_point( size=3) + 
  scale_shape_manual(values = c(3,4)) +
  facet_grid(. ~ Parameter ,scales = "free") +
  theme(legend.position="bottom")
```

### two-theta parameters

```{r plotCompareTwoThetaParams, fig.height=3}
bind_rows(.id = "implementation",
          new=map_df(twoThetaRes$twoThetaRes, ~ as_tibble(t(.x$par))) %>% 
            mutate( geneID = rownames(gene.data)),
          old=twoBSThetaMLparams %>% 
            mutate( geneID = rownames(gene.data))) %>% 
  gather(key = "Parameter",value = "Value", -implementation, -geneID) %>%
  myCompareParamPlot() +
  geom_vline(data=tibble(x=46.0517,Parameter="alpha"),aes(xintercept=x),color="red",linetype=2)
```

These parameter estimates looks way off, which is strange given that the likelihoods were so similar. Try calculating new log likihoods using the "old" parameter estimates and compare with the original old log likelihoods:

```{r plotCompareNewLLFromOldPar}

isThetaShiftEdge <- 1:Nedge(tree) %in% getEdgesFromMRCA(tree, tips = c("D","E"))
index.expand <- match(colnames(gene.data), tree$tip.label)

newLLfromOldParams <-
  twoBSThetaMLparams %>% 
  mutate( geneID = rownames(gene.data)) %>% 
  mutate( LL = pmap_dbl(.l = list(theta1=theta1,theta2=theta2,sigma2=sigma2,alpha=alpha,beta=beta,
                                  gene.data.row = lapply(geneID, function(i) gene.data[i, ])),
                        .f = logLikTwoTheta,
                        tree = tree, 
                        isThetaShiftEdge = isThetaShiftEdge,
                        index.expand = index.expand))

bind_rows( .id="implementation",
  old=tibble( twoThetaML=twoBSThetaMLs, geneID=rownames(gene.data)),
  reevaluated=tibble( twoThetaML=newLLfromOldParams$LL, geneID=newLLfromOldParams$geneID)) %>% 
  gather(key = "Parameter",value = "Value", -implementation, -geneID) %>%
  mutate( geneID = factor(geneID,levels=rev(unique(geneID)))) %>% 
  ggplot( aes(x=Value,y=geneID, color=implementation, shape=implementation)) + 
  geom_point( size=3) + 
  scale_shape_manual(values = c(3,4)) +
  facet_grid(. ~ Parameter ,scales = "free") +
  theme(legend.position="bottom")

```

The likelihoods for some genes are also far off. Most likely the original implementation does not return the correct parameters.

### Convergence time

```{r}
bind_rows( .id = "model",
  oneTheta = tibble(
    iterations = sapply(twoThetaRes$oneThetaRes, function(x) x$counts[1]),
    convergence = sapply(twoThetaRes$oneThetaRes, function(x) x$message ),
    geneID = rownames(gene.data)),
  twoTheta = tibble(
    iterations = sapply(twoThetaRes$twoThetaRes, function(x) x$counts[1]),
    convergence = sapply(twoThetaRes$twoThetaRes, function(x) x$message ),
    geneID = rownames(gene.data))
) %>% 
  mutate( geneID=factor(geneID,levels=rev(rownames(gene.data)))) %>% 
  ggplot() + 
  geom_point( aes(x=iterations,y=geneID, color=convergence) ) + 
  facet_wrap( ~model) + 
  theme(legend.position="bottom",legend.direction = "vertical")

```

