---
title: "Comparison with original EVE results (salmon20)"
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ape)

source('./scripts/eve-io.R')
source('./scripts/dvdt.R')
source("scripts/twoThetaTest.R")

library(tidyverse)
```


```{r Load data}
# simulated example data
tree <- read.tree("data/comparison/newFormat/salmonData/salmonidsPhylo.newick")
gene.data <- getExprData("data/comparison/newFormat/salmonData/salmon20.tsv")

# original EVE results
indivBetaMLparams <- read.table("data/comparison/oldEVE/results/indivBetaMLparams_salmon20.res", 
                                col.names = c("theta","sigma2","alpha","beta"))
sharedBetaMLparams <- read.table("data/comparison/oldEVE/results/sharedBetaMLparams_salmon20.res", 
                                 col.names = c("theta","sigma2","alpha","beta"))
betaTestLRTs <- scan("data/comparison/oldEVE/results/betaTestLRTs_salmon20.res")
sharedBetaMLs <- scan("data/comparison/oldEVE/results/sharedBetaMLs_salmon20.res")
indivBetaMLs <- scan("data/comparison/oldEVE/results/indivBetaMLs_salmon20.res")


# original EVE two-theta test results
oneThetaMLparams <- read.table("data/comparison/oldEVE/results/oneThetaMLparams_salmon20.res",
                                col.names = c("theta","sigma2","alpha","beta"))
twoBSThetaMLparams <-
  read.table("data/comparison/oldEVE/results/twoBSThetaMLparams_salmon20.res",
             col.names = c("theta","sigma2","alpha","beta")) %>%
  mutate(key=paste0("theta",rep(1:2,n()/2)), idx = rep(1:(n()/2),each=2)) %>%
  spread(key = key,value = theta) %>%
  arrange(idx) %>% select(-idx)
BSThetaTestLRTs <- scan("data/comparison/oldEVE/results/BSThetaTestLRTs_salmon20.res")
oneThetaMLs <- scan("data/comparison/oldEVE/results/oneThetaMLs_salmon20.res")
twoBSThetaMLs <- scan("data/comparison/oldEVE/results/twoBSThetaMLs_salmon20.res")

shiftSpecies <- c("Salp","Okis","Omyk","Ssal")
```

## Beta shared test

```{r Run beta-shared test, cache=TRUE, include=F}
BetaSharedRes <- betaSharedTest(tree, gene.data)
```

```{r defPlotFun}
myCompareParamPlot <- function(df){
  df %>% 
    mutate( Parameter = factor(Parameter,levels=unique(Parameter))) %>% 
    mutate( geneID = factor(geneID,levels=rev(unique(geneID)))) %>% 
    ggplot( aes(x=Value,y=geneID, color=implementation, shape=implementation)) + 
    geom_point( size=3) + 
    scale_shape_manual(values = c(3,4)) +
    facet_grid(. ~ Parameter ,scales = "free") +
    theme(legend.position="bottom")
}
```

### Beta shared estimate

* old: `r sharedBetaMLparams$beta[1]`
* new: `r BetaSharedRes$sharedBeta`

### LRTs

```{r plotLRT, fig.height=4}
bind_rows( .id="implementation",
  old=tibble( LRT=betaTestLRTs, geneID=rownames(gene.data)),
  new=tibble( LRT=BetaSharedRes$LRT, geneID=rownames(gene.data))) %>% 
  gather(key = "Parameter",value = "Value", -implementation, -geneID) %>%
  myCompareParamPlot() +
  geom_vline(data=tibble(x=qchisq(p=0.95,df=1),Parameter="LRT"),aes(xintercept=x),color="blue",linetype=3) +
  geom_vline(data=tibble(x=qchisq(p=0.05,df=1),Parameter="LRT"),aes(xintercept=x),color="blue",linetype=3)
```

Vertical lines are 0.05 and 0.95 of chi-squared distribution with 1 degree of freedom.

### MLs

```{r plotMLs, fig.height=4}
bind_rows( .id="implementation",
  old=tibble( indivBetaML=indivBetaMLs,sharedBetaML=sharedBetaMLs, geneID=rownames(gene.data)),
  new=tibble( indivBetaML=-map_dbl(BetaSharedRes$indivBetaRes,"value"),sharedBetaML=-map_dbl(BetaSharedRes$sharedBetaRes,"value"), geneID=rownames(gene.data))) %>% 
  gather(key = "Parameter",value = "Value", -implementation, -geneID) %>%
  mutate( geneID = factor(geneID,levels=rev(unique(geneID)))) %>% 
  ggplot( aes(x=Value,y=geneID, color=implementation, shape=implementation)) + 
  geom_point( size=3) + 
  scale_shape_manual(values = c(3,4)) +
  # scale_x_log10() +
  facet_grid(. ~ Parameter ,scales = "free") +
  theme(legend.position="bottom")
```



### individual beta parameters

```{r plotCompareIndivBetaParams, fig.height=4, message=F}
alphaMax <- -log(.01) / min(tree$edge.length[tree$edge[,2] <= Ntip(tree)])

bind_rows(.id = "implementation",
          new=map_df(BetaSharedRes$indivBetaRes, ~ as_tibble(t(.x$par))) %>% 
            mutate( geneID = rownames(gene.data)),
          old=indivBetaMLparams %>% 
            mutate( geneID = rownames(gene.data))) %>% 
  gather(key = "Parameter",value = "Value", -implementation, -geneID) %>%
  filter( Parameter != "theta") %>% 
  myCompareParamPlot() +
  geom_vline(data=tibble(x=alphaMax,Parameter="alpha"),aes(xintercept=x),color="red",linetype=2) + 
  scale_x_log10()


bind_rows(.id = "implementation",
          new=map_df(BetaSharedRes$indivBetaRes, ~ as_tibble(t(.x$par))) %>% 
            mutate( geneID = rownames(gene.data)),
          old=indivBetaMLparams %>% 
            mutate( geneID = rownames(gene.data))) %>% 
  gather(key = "Parameter",value = "Value", -implementation, -geneID) %>%
  filter( Parameter == "theta") %>% 
  myCompareParamPlot()
```

### shared beta parameters

```{r plotCompareSharedBetaParams, fig.height=4}
bind_rows(.id = "implementation",
          new=map_df(BetaSharedRes$sharedBetaRes, ~ as_tibble(t(.x$par))) %>% 
            mutate( geneID = rownames(gene.data)),
          old=sharedBetaMLparams %>% 
            mutate( geneID = rownames(gene.data))) %>% 
  gather(key = "Parameter",value = "Value", -implementation, -geneID) %>%
  filter( Parameter != "beta") %>% 
  filter( Parameter != "theta") %>% 
  myCompareParamPlot() +
  geom_vline(data=tibble(x=alphaMax,Parameter="alpha"),aes(xintercept=x),color="red",linetype=2) +
  scale_x_log10()


bind_rows(.id = "implementation",
          new=map_df(BetaSharedRes$sharedBetaRes, ~ as_tibble(t(.x$par))) %>% 
            mutate( geneID = rownames(gene.data)),
          old=sharedBetaMLparams %>% 
            mutate( geneID = rownames(gene.data))) %>% 
  gather(key = "Parameter",value = "Value", -implementation, -geneID) %>%
  filter( Parameter == "theta") %>% 
  myCompareParamPlot()

```

## Two-theta test results

```{r Run two-theta test, cache=TRUE, include=F}
twoThetaRes <- twoThetaTest(tree, gene.data,shiftSpecies = shiftSpecies)
```

### LRTs

```{r plotLRTtwoTheta, fig.height=3}
bind_rows( .id="implementation",
  old=tibble( LRT=BSThetaTestLRTs, geneID=rownames(gene.data)),
  new=tibble( LRT=twoThetaRes$LRT, geneID=rownames(gene.data))) %>% 
  gather(key = "Parameter",value = "Value", -implementation, -geneID) %>%
  myCompareParamPlot() +
  geom_vline(data=tibble(x=qchisq(p=0.95,df=1),Parameter="LRT"),aes(xintercept=x),color="blue",linetype=3) +
  geom_vline(data=tibble(x=qchisq(p=0.05,df=1),Parameter="LRT"),aes(xintercept=x),color="blue",linetype=3)
```

### MLs

```{r plotMLstwoTheta, fig.height=3}
bind_rows( .id="implementation",
  old=tibble( oneThetaML=oneThetaMLs,twoThetaML=twoBSThetaMLs, geneID=rownames(gene.data)),
  new=tibble( oneThetaML=-map_dbl(twoThetaRes$oneThetaRes,"value"),twoThetaML=-map_dbl(twoThetaRes$twoThetaRes,"value"), geneID=rownames(gene.data))) %>% 
  gather(key = "Parameter",value = "Value", -implementation, -geneID) %>%
  mutate( geneID = factor(geneID,levels=rev(unique(geneID)))) %>% 
  ggplot( aes(x=Value,y=geneID, color=implementation, shape=implementation)) + 
  geom_point( size=3) + 
  scale_shape_manual(values = c(3,4)) +
  facet_grid(. ~ Parameter ,scales = "free") +
  theme(legend.position="bottom")
```


### two-theta parameters

```{r plotCompareTwoThetaParams, fig.height=3}
bind_rows(.id = "implementation",
          new=map_df(twoThetaRes$twoThetaRes, ~ as_tibble(t(.x$par))) %>% 
            mutate( geneID = rownames(gene.data)),
          old=twoBSThetaMLparams %>% 
            mutate( geneID = rownames(gene.data))) %>% 
  gather(key = "Parameter",value = "Value", -implementation, -geneID) %>%
  filter( !grepl("theta",Parameter)) %>% 
  myCompareParamPlot() +
  geom_vline(data=tibble(x=alphaMax,Parameter="alpha"),aes(xintercept=x),color="red",linetype=2) + 
  scale_x_log10()


bind_rows(.id = "implementation",
          new=map_df(twoThetaRes$twoThetaRes, ~ as_tibble(t(.x$par))) %>% 
            mutate( geneID = rownames(gene.data)),
          old=twoBSThetaMLparams %>% 
            mutate( geneID = rownames(gene.data))) %>% 
  gather(key = "Parameter",value = "Value", -implementation, -geneID) %>%
  filter( grepl("theta",Parameter)) %>% 
  myCompareParamPlot()

```

Try calculating new log likihoods using the "old" parameter estimates and compare with the original old log likelihoods:
( )

```{r plotCompareNewLLFromOldPar}

isThetaShiftEdge <- 1:Nedge(tree) %in% getEdgesFromMRCA(tree, tips = shiftSpecies)
index.expand <- match(colnames(gene.data), tree$tip.label)

newLLfromOldParams <-
  twoBSThetaMLparams %>% 
  mutate( geneID = rownames(gene.data)) %>% 
  mutate( LL = pmap_dbl(.l = list(theta1=theta1,theta2=theta2,sigma2=sigma2,alpha=alpha,beta=beta,
                                  gene.data.row = lapply(geneID, function(i) gene.data[i, ])),
                        .f = logLikTwoTheta,
                        tree = tree, 
                        isThetaShiftEdge = isThetaShiftEdge,
                        index.expand = index.expand))

isThetaShiftEdge <- 1:Nedge(tree) %in% match(shiftSpecies,tree$tip.label)

newLLfromOldParamsNoInternal <-
  twoBSThetaMLparams %>% 
  mutate( geneID = rownames(gene.data)) %>% 
  mutate( LL = pmap_dbl(.l = list(theta1=theta1,theta2=theta2,sigma2=sigma2,alpha=alpha,beta=beta,
                                  gene.data.row = lapply(geneID, function(i) gene.data[i, ])),
                        .f = logLikTwoTheta,
                        tree = tree, 
                        isThetaShiftEdge = isThetaShiftEdge,
                        index.expand = index.expand))


bind_rows( .id="implementation",
  old=tibble( twoThetaML=twoBSThetaMLs, geneID=rownames(gene.data)),
  reevaluated=tibble( twoThetaML=newLLfromOldParams$LL, geneID=newLLfromOldParams$geneID),
  reevaluatedNoInternal=tibble( twoThetaML=newLLfromOldParamsNoInternal$LL, geneID=newLLfromOldParamsNoInternal$geneID)) %>% 
  gather(key = "Parameter",value = "Value", -implementation, -geneID) %>%
  mutate( geneID = factor(geneID,levels=rev(unique(geneID)))) %>% 
  ggplot( aes(x=Value,y=geneID, color=implementation, shape=implementation)) + 
  geom_point( size=3) + 
  scale_shape_manual(values = c(3,4,1)) +
  facet_grid(. ~ Parameter ,scales = "free") +
  theme(legend.position="bottom")

```

The likelihoods are almost identical despite the variation in the parameters, indicating instability in the parameters.

### Convergence time

```{r}
bind_rows( .id = "model",
  oneTheta = tibble(
    iterations = sapply(twoThetaRes$oneThetaRes, function(x) x$counts[1]),
    convergence = sapply(twoThetaRes$oneThetaRes, function(x) x$message ),
    geneID = rownames(gene.data)),
  twoTheta = tibble(
    iterations = sapply(twoThetaRes$twoThetaRes, function(x) x$counts[1]),
    convergence = sapply(twoThetaRes$twoThetaRes, function(x) x$message ),
    geneID = rownames(gene.data))
) %>% 
  mutate( geneID=factor(geneID,levels=rev(rownames(gene.data)))) %>% 
  ggplot() + 
  geom_point( aes(x=iterations,y=geneID, color=convergence) ) + 
  facet_wrap( ~model) + 
  theme(legend.position="bottom",legend.direction = "vertical")

```

