---
title: "Two-theta model analysis"
output:
  html_document: 
    toc: yes
    toc_float: yes
    code_folding: hide
editor_options: 
  chunk_output_type: console
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(ape)
library(ggiraph)

source("scripts/EVEcore.R")
source("scripts/eve-io.R")
source("scripts/twoThetaTest.R")
```

```{r loadData}
tree <- read.tree("data/comparison/newFormat/salmonData/salmonidsPhylo.newick")
gene.data <- getExprData("data/comparison/newFormat/salmonData/salmon20.tsv")

shiftSpecies <- c("Salp","Okis","Omyk","Ssal")
initParams <- initialParamsTwoTheta(gene.data,colSpecies = colnames(gene.data),shiftSpecies = shiftSpecies)

lowerBound <- c(theta1 = -99, theta2 = -99, sigma2 = 0.0001, alpha = 0.001, beta = 0.001)
upperBound <- c(theta1 =  99, theta2 =  99, sigma2 =   9999, alpha = 999  , beta = 99   )
```

```{r defFun}

parTransFuns <- list(
  none = list(
    trans = function(par){par},
    untrans = function(par){par}
  ),
  log = list(
    trans = function(par){
      with(as.list(par),
           c( theta1 = theta1, theta2 = theta2, 
              sigma2 = log(sigma2), alpha = log(alpha), beta = log(beta)))
    },
    untrans = function(par){
      with(as.list(par),
           c( theta1 = theta1, theta2 = theta2, 
              sigma2 = exp(sigma2), alpha = exp(alpha), beta = exp(beta)))
    }),
  rho = list(
    trans = function(par){
      with(as.list(par),
           c( theta1=theta1, theta2=theta2, alpha=alpha,
              rho=sigma2/(2*alpha), tau=beta*sigma2/(2*alpha)))
    },
    untrans = function(par){
      with(as.list(par),
           c( theta1=theta1, theta2=theta2, sigma2 = 2*rho*alpha,
              alpha=alpha, beta = tau/rho))
    })
)

fitTwoThetasTrans <- function( tree, shiftSpecies, colSpecies, gene.data.row, 
                               initParams, lowerBound, upperBound,
                               trans = c("none","log","rho")){
  
  transFun <- parTransFuns[[match.arg(trans)]]

  isThetaShiftEdge <- 1:Nedge(tree) %in% getEdgesFromMRCA(tree, tips = shiftSpecies)
  
  # match the column species with the phylogeny tip labels
  index.expand <- match(colSpecies, tree$tip.label)

  # function to optimize
  LLPerGeneTwoTheta <- function(par)
  {
    params <- transFun$untrans(par)
    ll <- logLikTwoTheta( theta1 = params[1], theta2 = params[2], sigma2 = params[3],
                          alpha = params[4], beta = params[5],
                          tree = tree, isThetaShiftEdge = isThetaShiftEdge, 
                          gene.data.row = gene.data.row, index.expand = index.expand)
    return(-ll)
  }
  
  # This doesn't seem to work..
  # if( trans == "rho"){
  #   # rho transform uses inverse of alpha, therefore swap upper and lower bound of alpha
  #   tmp <- lowerBound["alpha"]
  #   lowerBound["alpha"] <- upperBound["alpha"]
  #   upperBound["alpha"] <- tmp
  # }
    
  res <- optim( par = transFun$trans(initParams), fn = LLPerGeneTwoTheta, method = "L-BFGS-B",
                lower = transFun$trans(lowerBound), 
                upper = transFun$trans(upperBound))

  res$par <- transFun$untrans(res$par)
  
  return(res)
}

```

```{r fitAll, cache=T}
lapply(c("log","none") %>% setNames(.,.), function(trans){
  lapply(rownames(gene.data), function(geneID){
    fitTwoThetasTrans(tree = tree, shiftSpecies = shiftSpecies, colSpecies =colnames(gene.data), 
                      gene.data.row = gene.data[geneID, ], 
                      initParams = initParams[geneID, ], 
                      trans = trans,
                      lowerBound = lowerBound, upperBound = upperBound)
  })  
}) -> resAll

resTbl <- 
  lapply(resAll,function(resTrans){
    t(sapply(resTrans, function(res) res$par)) %>% 
    as.tibble() %>% 
    mutate( geneID=rownames(gene.data),
            ll = sapply(resTrans, function(res) -res$value),
            convergence = map_int(resTrans, "convergence"),
            iterations = sapply(resTrans, function(res) res$counts[1]))
  }) %>% 
  bind_rows(.id = "trans")
```

## log transform and bounds

### effect on iterations

```{r plotIterations}
resTbl %>% 
  mutate( geneID = factor(geneID,levels=rev(unique(geneID)))) %>% 
  mutate( convergence = convergence==0 ) %>% 
  ggplot( aes(x=geneID,y=iterations, fill=trans, color=convergence)) +
  geom_col(position = "dodge",width = 0.7) +
  scale_color_manual( values = c("green","black")) +
  coord_flip()
```

### parameter estimates

```{r plotPars}

resTbl %>%
  select(-theta1,-theta2, -iterations) %>% 
  group_by(geneID) %>%
  mutate( llDiff = ll - mean(ll)) %>% 
  ungroup() %>% 
  select(-ll) %>% 
  gather(key = "Parameter", value = "Value", -geneID, -trans, -convergence, -llDiff) %>%
  mutate( geneID = factor(geneID,levels=rev(unique(geneID)))) %>% 
  mutate( atUpperBound = Value > (upperBound[Parameter]*0.999)) %>%
  mutate( atLowerBound = Value < (lowerBound[Parameter]*1.001)) %>%
  mutate( atBounds = atUpperBound | atLowerBound) %>% 
  ggplot( aes(x=Value,y=geneID, color=llDiff, shape=trans)) + 
  geom_point( size=3) + 
  scale_shape_manual(values = c(3,4)) +
  scale_color_gradient2(low="blue",mid="black", high="red") +
  facet_grid(. ~ Parameter ,scales = "free") +
  theme(legend.position="bottom") +
  scale_x_log10() +
  geom_vline(data=tibble(x=upperBound[3:5],Parameter=names(upperBound[3:5])),aes(xintercept=x),
             color="red",linetype=3)

```


## overfitting alpha to the mean?

### Expected expression {.tabset}
```{r plotExp,results="asis"}
# plot the expression on Y and time since shift on the X. + Expected expression trend line.

# get time since shift
mrcaNode <- getMRCA(tree,shiftSpecies)
spcNodes <- match(shiftSpecies,tree$tip.label)
timeSinceShift <- setNames(dist.nodes(tree)[mrcaNode,spcNodes], shiftSpecies)

for( gene in rownames(gene.data)){

  cat("\n\n####",gene,"\n")
  
  with(as.list(resTbl %>% filter(geneID==gene, trans=="log")),{
    trendLine <- 
      tibble(timeSinceShift = seq(0,max(timeSinceShift),length.out = 100)) %>% 
      mutate( expLevel = theta2 + (theta1-theta2) * exp(-alpha * timeSinceShift))
    
    tibble(expLevel=gene.data[gene,], spc=colnames(gene.data)) %>% 
      mutate( timeSinceShift = ifelse(spc %in% shiftSpecies,timeSinceShift[spc],0) ) %>% 
      ggplot( aes( x = timeSinceShift, y = expLevel)) +
      geom_point() +
      geom_line( data=trendLine, color="red",linetype=2) +
      geom_hline(yintercept=theta1, color="blue",linetype=3) +
      annotate( "text", label="theta1",x=0.02,y=theta1, color="blue", vjust=-0.5) +
      geom_hline(yintercept=theta2, color="blue",linetype=3) + 
      annotate( "text", label="theta2",x=0.02,y=theta2, color="blue", vjust=-0.5) +
      ggtitle(paste0(gene,"  sigma2 = ",signif(sigma2,3),"  alpha = ",signif(alpha,3), "  beta = ", signif(beta,3)))
  }) -> g
  print(g)
  
}

```

## Alpha and Ssal expression

Beacause of the phylogeny where Ssal has a significantly shorter branch than the other species and the decision to define the theta shift at the most recent common ancestor of the salmonids

```{r relSsalShiftPlot}
colSpeciesIndices <- split(1:ncol(gene.data), f = colnames(gene.data))
species.mean <- sapply(colSpeciesIndices, function(i){ rowMeans(gene.data[,i]) })

tibble(
  geneID = rownames(gene.data),
  preShift = rowMeans(species.mean[ ,c("Drer", "Eluc", "Olat")]),
  postShift = rowMeans(species.mean[ ,c("Salp", "Okis", "Omyk")]),
  Ssal = species.mean[ ,"Ssal"]
) %>% 
  mutate( relSsalShift = (Ssal-preShift) / (postShift-preShift)) %>% 
  left_join( resTbl %>% filter(trans=="log") %>% select(geneID,alpha), by="geneID") %>% 
  ggplot( aes(x=relSsalShift, y=alpha, color=alpha)) +
  scale_color_gradientn(colours = c("blue","black","red"),trans = "log", breaks = c(10^(-1:3))) +
  geom_point_interactive( aes(tooltip=geneID, data_id = geneID)) + 
  scale_y_log10() -> g

girafe(code = print(g))
```

## Between-/Within-species variance
```{r betweenSpeciesVarPlot}

# calculate expected mean based on parameters
expLevel <- species.mean*0
for( gene in rownames(species.mean)){
  with(as.list(resTbl %>% filter(geneID==gene, trans=="log")),{
    for( spc in colnames(species.mean)){
      dt <- ifelse( spc %in% names(timeSinceShift),timeSinceShift,0)
      expLevel[gene,spc] <<- theta2 + (theta1-theta2) * exp(-alpha * dt)
    }
  })
}

# subtract the expected mean before calculating between species mean
species.mean.adjusted <- species.mean - expLevel

tibble(
  geneID = rownames(gene.data),
  betweenSpeciesVar = apply(species.mean.adjusted,1,var)
) %>% 
  left_join( resTbl %>% filter(trans=="log"), by="geneID") %>% 
  ggplot( aes(x=betweenSpeciesVar, y=sigma2/(2*alpha), color=alpha)) +
  scale_color_gradientn(colours = c("blue","black","red"),trans = "log", breaks = c(10^(-1:3))) +
  geom_point() +
  geom_point_interactive( aes(tooltip=geneID, data_id = geneID)) + 
  geom_abline( slope=1, intercept=0, color="red", linetype=2) +
  scale_x_log10() +
  scale_y_log10() -> g

girafe(code = print(g))  
```

```{r withinSpeciesVarPlot}
tibble(
  geneID = rownames(gene.data),
  withinSpeciesVar = rowMeans(sapply(colSpeciesIndices, function(i){ apply(gene.data[,i],1,var) }))
) %>% 
  left_join( resTbl %>% filter(trans=="log"), by="geneID") %>% 
  ggplot( aes(x=withinSpeciesVar, y=beta*sigma2/(2*alpha), color=alpha)) +
  scale_color_gradientn(colours = c("blue","black","red"),trans = "log", breaks = c(10^(-1:3))) +
  geom_point() +
  geom_point_interactive( aes(tooltip=geneID, data_id = geneID)) + 
  geom_abline( slope=1, intercept=0, color="red", linetype=2) +
  scale_x_log10() +
  scale_y_log10() -> g

girafe(code = print(g))  
```


